---
title: "Multivariate analyses"
bibliography: references.bib
link-citations: true
date: "Last compiled on `r format(Sys.time(), '%B, %Y')`"
output: 
  html_document:
    css: tweaks.css
    toc:  true
    toc_float: true
    number_sections: false
    toc_depth: 2
    code_folding: show
    code_download: yes
---

```{r, globalsettings, echo=FALSE, warning=FALSE, results='hide',message=FALSE}
library(knitr)
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
opts_chunk$set(tidy.opts=list(width.cutoff=100),tidy=TRUE, warning = FALSE, message = FALSE,comment = "#>", cache=TRUE, class.source=c("test"), class.output=c("test3"))
options(width = 100)
rgl::setupKnitr()


colorize <- function(x, color) {sprintf("<span style='color: %s;'>%s</span>", color, x) }
```


```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'))
#klippy::klippy(color = 'darkred')
#klippy::klippy(tooltip_message = 'Click to copy', tooltip_success = 'Done')
```



---  
  
# Getting started

To copy the code, click the button in the upper right corner of the code-chunks.

## clean up

```{r, eval=FALSE, results='hide'}
rm(list=ls())
gc()
```

<br>

## general custom functions

- `fpackage.check`: Check if packages are installed (and install if not) in R
- `fsave`: Function to save data with time stamp in correct directory
- `fload`: Function to load R-objects under new names
- `fshowdf`: Print objects (`tibble` / `data.frame`) nicely on screen in `.Rmd`.
- `ffit`: fit a series of (here, generalized linear mixed-effects) models 

```{r}
fpackage.check <- function(packages) {
    lapply(packages, FUN = function(x) {
        if (!require(x, character.only = TRUE)) {
            install.packages(x, dependencies = TRUE)
            library(x, character.only = TRUE)
        }
    })
}

fsave <- function(x, file, location = "./data/processed/", ...) {
    if (!dir.exists(location))
        dir.create(location)
    datename <- substr(gsub("[:-]", "", Sys.time()), 1, 8)
    totalname <- paste(location, datename, file, sep = "")
    print(paste("SAVED: ", totalname, sep = ""))
    save(x, file = totalname)
}

fload  <- function(fileName){
  load(fileName)
  get(ls()[ls() != "fileName"])
}

fshowdf <- function(x, digits = 2, ...) {
    knitr::kable(x, digits = digits, "html", ...) %>%
        kableExtra::kable_styling(bootstrap_options = c("striped", "hover")) %>%
        kableExtra::scroll_box(width = "100%", height = "300px")
}

ffit <- function(formula, data) {
  tryCatch({
    model <- lme4::glmer(formula, data = data, family = binomial(link = "logit"),
                   control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5)))
    cat("Fitting model:", as.character(formula), "\n")
    summary(model)
    cat("\n")
    return(model)
  }, error = function(e) {
    cat("Error fitting model:", as.character(formula), "\n")
    cat("Error message:", conditionMessage(e), "\n")
    return(NULL)
  })
}
```

<br>

## necessary packages

```{r, results='hide', message=FALSE, warning=FALSE}
packages = c("tidyverse", "ggplot2", "GenBinomApps")
fpackage.check(packages)
rm(packages)
``` 

<br>

## load data-set

Load the replicated data-set. To load the file, adjust the filename in the following code so that it matches the most recent version of the `.RDa` file you have in your `./data/processed/` folder.

You may also obtain them by downloading: `r xfun::embed_file("./data_shared/data_nested.RDa")`


```{r}
#list files in processed data folder
list.files("./data/processed/")

#get todays date:
today <- gsub("-", "", Sys.Date() - 1)

#use fload
df <- fload(paste0("./data/processed/", today, "data_nested.RDa"))
```

<br>

## last alterations

- standardize embeddedness in other network layers
- proximity levels

```{r}
df$embed.ext <- df$embed.ext/3
df$proximity <- factor(df$proximity, levels = c("far","close","roommate"))



```

<br>

# Descriptives

**who maintains educationally (dis)similar friends and confidants, and does this depend on one's mobility trajectory?**

```{r}
# function to calculate proportion self-similar and dis-similar alters that is maintained (and their ratio), for a given group (e.g., uni), tie type (e.g, friend), period (e.g., first period), cohort (e.g., first years), and mobility trajectory (e.g., upwardly mobile)

fprops <- function(df, ego_educ = NULL, tie = NULL, mobility = NULL, period = NULL, year = NULL) {
  subset_df <- df
  # apply filters if specified
  if (!is.null(ego_educ)) {
    subset_df <- subset_df[subset_df$ego_educ == ego_educ, ]
  }
  if (!is.null(tie)) {
    subset_df <- subset_df[subset_df$tie == tie, ]
  }
  if (!is.null(mobility)) {
    subset_df <- subset_df[subset_df$mobility == mobility, ]
  }
  if (!is.null(period)) {
    subset_df <- subset_df[subset_df$period == period, ]
  }
  if (!is.null(year)) {
    subset_df <- subset_df[subset_df$study.year == year, ]
  }
  
  # means for sim_educ == 1 and sim_educ == 0 and the ratio
  mean_sim1 <- mean(subset_df[subset_df$sim_educ == 1, ]$Y, na.rm = TRUE)
  n1 <- nrow(subset_df[subset_df$sim_educ == 1,])
  mean_sim0 <- mean(subset_df[subset_df$sim_educ == 0, ]$Y, na.rm = TRUE)
  n0 <- nrow(subset_df[subset_df$sim_educ == 0,])
  mean_ratio <- mean_sim1 / mean_sim0
  
  return(list(mean_sim1 = mean_sim1,
              mean_sim0 = mean_sim0,
              mean_ratio = mean_ratio,
              n1 = n1, 
              n0 = n0))
}

fprops(df)

# university students drop non-university friends more often:
fprops(df, ego_educ = "7", tie = "Friend")
fprops(df, ego_educ = "6", tie = "Friend")

# stronger in first period
fprops(df, ego_educ = "7", tie = "Friend", period = "w1 -> w2")
fprops(df, ego_educ = "7", tie = "Friend", period = "w2 -> w3")

# also for confidants:
fprops(df, ego_educ = "7", tie = "Confidant")
fprops(df, ego_educ = "6", tie = "Confidant")

# depending on (intergenerational) mobility trajectory?
fprops(df, ego_educ = "7", tie = "Friend", mobility = "stable", period = "w1 -> w2")
fprops(df, ego_educ = "7", tie = "Friend", mobility = "stable", period = "w2 -> w3")

fprops(df, ego_educ = "7", tie = "Friend", mobility = "ascending", period = "w1 -> w2")
fprops(df, ego_educ = "7", tie = "Friend", mobility = "ascending", period = "w2 -> w3")
``` 

<br>

----


# References

