---
title: "Multivariate analyses"
bibliography: references.bib
link-citations: true
date: "Last compiled on `r format(Sys.time(), '%B, %Y')`"
output: 
  html_document:
    css: tweaks.css
    toc:  true
    toc_float: true
    number_sections: false
    toc_depth: 2
    code_folding: show
    code_download: yes
---

```{r, globalsettings, echo=FALSE, warning=FALSE, results='hide',message=FALSE}
library(knitr)
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
opts_chunk$set(tidy.opts=list(width.cutoff=100),tidy=TRUE, warning = FALSE, message = FALSE,comment = "#>", cache=TRUE, class.source=c("test"), class.output=c("test3"))
options(width = 100)
rgl::setupKnitr()


colorize <- function(x, color) {sprintf("<span style='color: %s;'>%s</span>", color, x) }
```


```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'))
#klippy::klippy(color = 'darkred')
#klippy::klippy(tooltip_message = 'Click to copy', tooltip_success = 'Done')
```



---  
  
# Getting started

To copy the code, click the button in the upper right corner of the code-chunks.

## clean up

```{r, eval=FALSE, results='hide'}
rm(list=ls())
gc()
```

<br>

## general custom functions

- `fpackage.check`: Check if packages are installed (and install if not) in R
- `fsave`: Function to save data with time stamp in correct directory
- `fload`: Function to load R-objects under new names
- `fshowdf`: Print objects (`tibble` / `data.frame`) nicely on screen in `.Rmd`.
- `ffit`: fit a series of (here, generalized linear mixed-effects) models 

```{r}
fpackage.check <- function(packages) {
    lapply(packages, FUN = function(x) {
        if (!require(x, character.only = TRUE)) {
            install.packages(x, dependencies = TRUE)
            library(x, character.only = TRUE)
        }
    })
}

fsave <- function(x, file, location = "./data/processed/", ...) {
    if (!dir.exists(location))
        dir.create(location)
    datename <- substr(gsub("[:-]", "", Sys.time()), 1, 8)
    totalname <- paste(location, datename, file, sep = "")
    print(paste("SAVED: ", totalname, sep = ""))
    save(x, file = totalname)
}

fload  <- function(fileName){
  load(fileName)
  get(ls()[ls() != "fileName"])
}

fshowdf <- function(x, digits = 2, ...) {
    knitr::kable(x, digits = digits, "html", ...) %>%
        kableExtra::kable_styling(bootstrap_options = c("striped", "hover")) %>%
        kableExtra::scroll_box(width = "100%", height = "300px")
}

ffit <- function(formula, data, link = "logit") {
  tryCatch({
    model <- lme4::glmer(formula, data = data, family = binomial(link = link),
                   control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5)))
    cat("Fitting model:", as.character(formula), "\n")
    summary(model)
    cat("\n")
    return(model)
  }, error = function(e) {
    cat("Error fitting model:", as.character(formula), "\n")
    cat("Error message:", conditionMessage(e), "\n")
    return(NULL)
  })
}
```

<br>

## necessary packages

```{r, results='hide', message=FALSE, warning=FALSE}
packages = c("tidyverse", "ggplot2", "lme4", "lmtest")
fpackage.check(packages)
rm(packages)
``` 

<br>

## load data-set

Load the replicated data-set. To load the file, adjust the filename in the following code so that it matches the most recent version of the `.RDa` file you have in your `./data/processed/` folder.

You may also obtain them by downloading: `r xfun::embed_file("./data_shared/data.RDa")`


```{r}
#list files in processed data folder
list.files("./data/processed/")

#get todays date:
today <- gsub("-", "", Sys.Date() -1 )

#use fload
#df <- fload(paste0("./data/processed/", today, "data_nested.RDa")) #this is a dataset containing ties (e.g., sport/friend) within dyads

df <- fload(paste0("./data/processed/", today, "data.RDa")) # this dataset is at the "alter-level", focusing only on friends/confidants (treating ties as maintained as they re-appear in the friend/confidant name generator)

# we have data on ...
length(unique(df$ego)) # 493 unique egos
length(unique(paste0(df$ego, "x", df$alterid))) # 2450 unique alters (friend/confidant)
nrow(df) * prop.table(table(df$period)) # who were observed 3256 times across observations
```

<br>

---

# Descriptives {.tabset .tabset-fade}

## ego

```{r, class.source = 'fold-hide'}
egos <- df %>%
  group_by(ego) %>%
  summarise(
    age = first(ego_age),
    female = first(ego_female),
    male = 1 - first(ego_female),
    ego_educ = first(ego_educ),
    study_year = first(study.year),
    father_educ = first(father_rank),
    mother_educ = first(mother_rank),
    mobility = first(mobility),
    .groups = "drop"
  ) %>%
  mutate(
    uni = ifelse(ego_educ == 7, 1, 0),
    hbo = ifelse(ego_educ == 6, 1, 0),
    first_year = ifelse(study_year == 1, 1, 0),
    second_year = ifelse(study_year == 2, 1, 0),
    third_year = ifelse(study_year == 3, 1, 0),
    mobility_descending = ifelse(mobility == "descending", 1, 0),
    mobility_stable = ifelse(mobility == "stable", 1, 0),
    mobility_ascending = ifelse(mobility == "ascending", 1, 0)
  ) %>%
  select(-ego, -ego_educ, -mobility, -study_year)

fshowdf(psych::describe(egos)[-c(1,6,7,11,12, 13)], caption="ego-level descriptives")
```

## alter

```{r, class.source = 'fold-hide'}
alters <- df %>%
  mutate(
    alter_key = paste0(ego, "x", alterid)
  ) %>%
  group_by(alter_key) %>%
  summarise(
    female = first(alter_female),
    male = 1- first(alter_female),
    age  = first(as.numeric(alter_age)),
    alter_educ     = first(alter_educ),
    same_gender    = first(same_gender),
    dif_age        = first(dif_age),
    sim_educ       = first(sim_educ),
    
    # create dummies for alter_educ
    primary = ifelse(first(alter_educ) == 1, 1, 0),
    vmbo_mavo = ifelse(first(alter_educ) == 2, 1, 0),
    mbo      = ifelse(first(alter_educ) == 3, 1, 0),
    havo     = ifelse(first(alter_educ) == 4, 1, 0),
    vwo_gymnasium = ifelse(first(alter_educ) == 5, 1, 0),
    hbo      = ifelse(first(alter_educ) == 6, 1, 0),
    uni       = ifelse(first(alter_educ) == 7, 1, 0),
    #proximity
    close    = ifelse(first(proximity) == "close", 1, 0),
    roommate = ifelse(first(proximity) == "roommate", 1, 0),
    far = ifelse(first(proximity) == "far", 1, 0),
    
    duration = first(duration),
    .groups = "drop"
  ) %>%
  select(-alter_key, -alter_educ)

fshowdf(psych::describe(alters)[-c(1,6,7,11,12, 13)], caption="alter-level descriptives")
```

## observations

```{r, class.source = 'fold-hide'}
obs <- df %>%
  mutate(
    alter_key = paste0(ego, "x", alterid)
  ) %>%
  select(
    alter_key, closeness.t, frequency.t, friend, confidant,
    multiplex, sport, study, size, embed, maintained, period
  ) %>%
  mutate(
    period_1 = ifelse(period == "1", 1, 0),
    period_2_maintained = ifelse(period == "2 (maintained)", 1, 0),
    period_2_created = ifelse(period == "2 (created)", 1, 0)
  ) %>%
  select(-ego, -alter_key, -period)

fshowdf(psych::describe(obs[,-1])[-c(1,6,7,11,12, 13)], caption="observation-level descriptives")
``` 


# {.unlisted .unnumbered}

---

## Homophily

**Who maintains educationally (dis)similar friends and confidants, and does this depend on one's mobility trajectory?**

calculate the intra- and inter-group tie maintenance probabilities, and their ratio, for specific subgroups:

```{r}
# function to calculate proportion self-similar and dis-similar alters that is maintained (and their ratio), for a given group (e.g., uni), period (e.g., first period), cohort (e.g., first years), and mobility trajectory (e.g., upwardly mobile)

fprops <- function(df, ego_educ = NULL, mobility = NULL, period = NULL, year = NULL) {
  subset_df <- df
  # apply filters if specified
  if (!is.null(ego_educ)) {
    subset_df <- subset_df[subset_df$ego_educ == ego_educ, ]
  }
  if (!is.null(mobility)) {
    subset_df <- subset_df[subset_df$mobility == mobility, ]
  }
  if (!is.null(period)) {
    subset_df <- subset_df[subset_df$period == period, ]
  }
  if (!is.null(year)) {
    subset_df <- subset_df[subset_df$study.year == year, ]
  }

  # means for sim_educ == 1 and sim_educ == 0 and the ratio
  mean_sim1 <- mean(subset_df[subset_df$sim_educ == 1, ]$maintained, na.rm = TRUE)
  n1 <- nrow(subset_df[subset_df$sim_educ == 1,])
  mean_sim0 <- mean(subset_df[subset_df$sim_educ == 0, ]$maintained, na.rm = TRUE)
  n0 <- nrow(subset_df[subset_df$sim_educ == 0,])
  mean_ratio <- mean_sim1 / mean_sim0
  
  return(list(mean_sim1 = mean_sim1,
              mean_sim0 = mean_sim0,
              mean_ratio = mean_ratio,
              n1 = n1, 
              n0 = n0,
              obs = n1 + n0,
              alters = length(unique(paste0(subset_df$ego, "x", subset_df$alterid))),
              egos = length(unique(subset_df$ego))))
}
```

```{r}
# first consider all observations (pooled)
fprops(df)

# university students have a bias towards dropping dissimilar alters:
fprops(df, ego_educ = "7")

# slightly weaker among hbo students (but low N ...)
fprops(df, ego_educ = "6")

# excluding hbo students' university friends strengthens the homophily bias, suggesting that the bias reflects not just a preference for similarity, but possibly an avoidance of 'downward' ties:
df %>%
  filter(alter_educ != 7 & ego_educ == 6) %>%
  fprops()

# does bias depend on the period in which tie change is observed?
fprops(df, ego_educ = "7", period = "1") # strongest in the 1st period
fprops(df, ego_educ = "7", period = "2 (maintained)") # in part because the second period is (a) shorter and (b) includes alters who already "survived" one round--the effect of dissimilarity plausibly weakens when ego and alter keep in touch.
fprops(df, ego_educ = "7", period = "2 (created)") # consistent with this idea, especially among newly formed ties the similarity bias is strong! 

# does bias depend on (intergenerational) mobility trajectory? NOT if we compare stable and ascending among the pool of university students...
fprops(df, ego_educ = "7", mobility = "stable")
fprops(df, ego_educ = "7", mobility = "ascending")

# among hbo students?
# here we see that bias towards dropping educationally dissimilar friends is strongest among ascending students, followed by--in order--stable and descending students. Descending hbo-students actually maintain dissimilar friends more often!
fprops(df, ego_educ = "6", mobility = "stable")
fprops(df, ego_educ = "6", mobility = "ascending")
fprops(df, ego_educ = "6", mobility = "descending")

# what if we exclude hbo students university friends?
# now the stable have the strongest ingroup bias (or bias against downward ties)
fprops(df[df$ego_educ=="6" & df$alter_educ!=7, ], mobility = "stable")
fprops(df[df$ego_educ=="6" & df$alter_educ!=7, ], mobility = "ascending")
fprops(df[df$ego_educ=="6" & df$alter_educ!=7, ], mobility = "descending") # so the aversion towards similar others is not due to a tendency to maintain university ('higher') friends
``` 

---

# Multivariate analyses

## random intercept logit model

```{r}
# focus on uni students for now
df2 <- df[df$ego_educ == "7", ]

# flat model (assuming no nesting of alters within egos)
m0 <- glm(maintained ~ 1, data = df2, family = binomial(link = "logit"))
#summary(m0)

# add random ego-level intercept
# predicting the odds of an alter being dropped from the network with no independent variables
m01 <- glmer(maintained ~ 1 + (1 | ego), data = df2, family = binomial(link = "logit"), control = glmerControl(optimizer="bobyqa"))
#summary(m01)

# retrieve variance components
varcomp <- VarCorr(m01)
var2 <- varcomp$ego[1]
var1 <- (pi^2)/3
vpc2 <- (var2)/(var1+var2)
vpc2
# the model shows that the proportion of variance in the outcome (tie maintenance) explained by between-respondent variation is .074.

# perform likelihood ratio test for differences in models
lrtest(m0, m01)
# this shows that the p-value for the null hypothesis that the variance at the ego-level is 0 is .000.

# add fixed effects for period
m02 <- glmer(maintained ~ 1 + (1 | ego) + period, data = df2, family = binomial(link = "logit"), control = glmerControl(optimizer="bobyqa"))
summary(m02)

# now let's introduce sets of variables: ego; alter; network; similarity; moderating role of mobility trajectory.
# first define some last variables: 
df2$sportstudy <- ifelse(df2$sport + df2$study > 1, 1, 0) # combine additional network layers
df2$close <- ifelse(df2$proximity == "close", 1, 0)
df2$roommate <- ifelse(df2$proximity == "roommate", 1, 0)
df2$firstgen <- ifelse(df2$mobility == "ascending", 1, 0)

formula <- list(
  #0. null model
  maintained ~ 1 + (1 | ego),
  #1. incl. fixed effects of period*status
  maintained ~ 1 + (1 | ego) + period,
  #2. ego 
  maintained ~ 1 + (1 | ego) + period + ego_female + as.factor(study.year),
  #3. alter
  maintained ~ 1 + (1 | ego) + period + ego_female + as.factor(study.year) + 
              confidant + multiplex + sportstudy + scale(duration) + close + roommate + frequency.t + closeness.t,
  #4. network
  maintained ~ 1 + (1 | ego) + period + ego_female + as.factor(study.year) + 
              confidant + multiplex + sportstudy + scale(duration) + close + roommate + frequency.t + closeness.t + size,
  #5. dyadic similarity
  maintained ~ 1 + (1 | ego) + period + ego_female + as.factor(study.year) + confidant + multiplex + sportstudy + scale(duration) + close + roommate + frequency.t + closeness.t + size + sim_educ + same_gender + dif_age,
  
  #6. interaction between same education and first-generation student
    maintained ~ 1 + (1 | ego) + period + ego_female + as.factor(study.year) + confidant + multiplex + sportstudy + scale(duration) + close + roommate + frequency.t + closeness.t + size + sim_educ + same_gender + dif_age + firstgen*sim_educ
  )

#estimate using ffit
ans <- lapply(formula, ffit, data = df2)

#use likelihood ratio test to compare models
do.call(lrtest, ans)
``` 
```{r, eval=FALSE, echo=FALSE}
texreg::htmlreg(ans,
        file="./results/coeftab.html",
        caption="Results of models predicting tie maintenance at t+1 (1=yes, 0=no)", caption.above = TRUE, 
        custom.model.names = c("Null model", "Period effects", "Ego controls", "Alter controls", "Network controls", "Dyadic similarity", "Interaction"),
        custom.coef.names = c("(Intercept)", 
                              "Period 2 (created)",
                              "Period 2 (maintained)",
                              "Female", "Second-year (ref. = first-year", "Third-year or higher",
                              "Confidant (ref. = friend)", "Multiplex", "Sport / study", "Years known", "Same municipality (ref. = farther away)", "Same house", "Communication frequency", "Emotional closeness", "Network size", "Same education", "Same gender", "Age difference", "First-gen student (ref. = continuing-gen)", "Same education:first-gen student"),
        digits=2, single.row = TRUE
        )
```

```{r, echo = FALSE}
htmltools::tags$div(
  style = "height: 600px; overflow-y: scroll;",
  htmltools::includeHTML("./results/coeftab.html")
)
```


<br>

## marginal (interaction) effects

Calculate and visualize (average / distribution of) marginal (interaction) effects.

### marginal effect of same education

```{r, fig.width=10}
# 1. define datasets
# educational similarity:
df2sim1 <- df2sim0 <- df2
df2sim1$sim_educ <- 1
df2sim0$sim_educ <- 0

# 2. interaction with first-gen
df2asc1sim1 <- df2asc1sim0 <- df2asc0sim1 <- df2asc0sim0 <- df2
{
  df2asc1sim1$firstgen <- 1
  df2asc1sim1$sim_educ <- 1  
  df2asc0sim1$firstgen <- 0
  df2asc0sim1$sim_educ <- 1  
  df2asc1sim0$firstgen <- 1
  df2asc1sim0$sim_educ <- 0
  df2asc0sim0$firstgen <- 0
  df2asc0sim0$sim_educ <- 0
}

m4 <- ans[[6]]
m5 <- ans[[7]]

# marginal effect of similarity:
#mean(lme4:::predict.merMod(m4, type = "response", re.form = NULL, newdata = df2sim1) - lme4:::predict.merMod(m4, type = "response", re.form = NULL, newdata = df2sim0))
me_sim <- lme4:::predict.merMod(m5, type = "response", re.form = NULL, newdata = df2sim1) - lme4:::predict.merMod(m5, type = "response", re.form = NULL, newdata = df2sim0)

# plot
dens <- ggplot(data.frame(me_sim), aes(x = me_sim)) +
  geom_density(fill = "lightblue", alpha = 0.5) +
  geom_vline(aes(xintercept = mean(me_sim)), color = "red", linetype = "dashed") +
  annotate("text", x = mean(me_sim), y = max(density(me_sim)$y), vjust = 3, hjust = -0.1, label = paste0("AME = ", round(mean(me_sim), 3)), color = "red", size = 3.5) +
  labs(title = "Density curve of marginal effects",
       x = "Marginal effect (first difference)",
       y = "Density") +
  theme_minimal() +
  xlim(0, NA)

# calculate predicted probabilities, include random effects.
preds <- lme4:::predict.merMod(m5, type = "response", re.form = NULL, newdata = df2) 

dens2 <- ggplot(data.frame(me_sim, preds), aes(x = preds, y = me_sim)) +
  geom_point(alpha = 0.4, color = "forestgreen") +
   geom_smooth(method = "loess", se = TRUE, color = "darkorange") +
  labs(
    title = "Marginal effect vs. predicted probability",
    x = "Predicted probability",
    y = "Marginal effect"
  ) +
  theme_minimal()

ggpubr::ggarrange(dens,dens2, labels = c("(A)", "(B)"))
```

On average, across all university students in our dataset and their observed characteristics, having a friend who also attends university increases the probability of maintaining the relationship by approximately 6 percentage points.

<br>

### interaction with first-gen student

```{r, fig.width=10}
# marginal interaction effect:
p11 <- lme4:::predict.merMod(m5, type = "response", re.form = NULL, newdata = df2asc1sim1)
p10 <- lme4:::predict.merMod(m5, type = "response", re.form = NULL, newdata = df2asc1sim0)
p01 <- lme4:::predict.merMod(m5, type = "response", re.form = NULL, newdata = df2asc0sim1)
p00 <- lme4:::predict.merMod(m5, type = "response", re.form = NULL, newdata = df2asc0sim0)

mie_sim_asc <- (p11 - p10) - (p01 - p00)

#represents how the marginal effect of same education changes when going from continuing-gen to first-gen student:
dens <- ggplot(data.frame( mie_sim_asc), aes(x = mie_sim_asc)) +
  geom_density(fill = "lightblue", alpha = 0.5) +
  geom_vline(aes(xintercept = mean(mie_sim_asc)), color = "darkblue", linetype = "dashed") +
  annotate(
    "text",
    x = mean(mie_sim_asc),
    y = max(density(mie_sim_asc)$y),
    vjust = 3, hjust = -0.1,
    label = paste0("AMIE = ", round(mean(mie_sim_asc), 3)),
    color = "darkblue", size = 3.5
  ) +
  labs(
    title = "Density curve of marginal interaction effects",
    x = "Marginal interaction effect (difference-in-differences)",
    y = "Density"
  ) +
  theme_minimal() +
  xlim(min(mie_sim_asc), max(mie_sim_asc))

dens2 <- ggplot(data.frame(mie_sim_asc, preds), aes(x = preds, y = mie_sim_asc)) +
  geom_point(alpha = 0.4, color = "forestgreen") +
  geom_smooth(method = "loess", se = TRUE, color = "darkorange") +
  labs(
    title = "Marginal interaction effect vs. predicted probability",
    x = "Predicted probability",
    y = "Marginal interaction effect"
  ) +
  theme_minimal()

ggpubr::ggarrange(dens,dens2, labels = c("(A)", "(B)"))
``` 

The average marginal effect is approximately 2.2 percentage points smaller for first-generation students compared to their continuing-generation counterparts--on average, across all university students in our sample-- indicating that educational similarity matters less for tie maintenance among first-gen students. While modest in absolute terms, in relative terms, this reduction represents more than 1/3 of the average marginal effect.

<br>

```{r, fig.width=10}
# predictions based on observed covariates
df2$pred <- preds

# predictions with covariates at the sample mean (continuous) or reference level
df3 <- df2
#all_vars <- names(fixef(m5))[-1]  # remove intercept

df3$period = "1"
df3$ego_female = 1
df3$study.year = 1
df3$confidant = 0
df3$multiplex = 0
df3$sportstudy = 0
df3$duration <- mean(df3$duration)
df3$close <- 0
df3$roommate <- 0
df3$frequency.t <- mean(df3$frequency.t)
df3$closeness.t <- mean(df3$closeness.t)
df3$size <- mean(df3$size)
df3$dif_age <- mean(df3$dif_age)
df3$same_gender <- 0

df3$pred <- lme4:::predict.merMod(m5, type = "response", re.form = NULL, newdata = df3)

obs <- ggplot(df2, aes(x = factor(sim_educ, levels = c(0, 1)), y = pred, color = factor(firstgen))) +
  geom_line(aes(group = firstgen), size = 1.5) +  
  geom_point(size = 3, alpha = 0.1) +  
  labs(
    x = "Same education",
    y = "Predicted probability",
    color = "First-generation student",
    title = "Predicted probabilities based on similarity and mobility",
    subtitle = "using observed covariates"
  ) +
  theme_minimal()

mea <- ggplot(df3, aes(x = factor(sim_educ, levels = c(0, 1)), y = pred, color = factor(firstgen))) +
  geom_line(aes(group = firstgen), size = 1.5) + 
  geom_point(size = 3, alpha = 0.1) + 
  labs(
    x = "Same education",
    y = "Predicted probability",
    color = "First-generation student",
    title = "Predicted probabilities based on similarity and mobility",    
    subtitle = "covariates set at the mean / reference level"
  ) +
  theme_minimal()

ggpubr::ggarrange(obs,mea, labels = c("(A)", "(B)"), common.legend = TRUE, legend="bottom")
```




